---
title: "FinalP1"
author: "Jonah Smith"
date: "2025-04-09"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Load Libraries

```{r, load_tidyverse}
library(tidyverse)
```

##Read Data

```{r, read_final_data}
df <- readr::read_csv("paint_project_train_data.csv", col_names = TRUE)
glimpse(df)
```

#Part i: Exploration

• Visualize the distributions of variables in the data set.

• Counts for categorical variables.

```{r}
df %>% ggplot(aes(x = Lightness)) + geom_bar(stat = "count") + labs(title = "Counts of the lightness variable")

df %>% ggplot(aes(x = Saturation)) + geom_bar(stat = "count") + labs(title = "Counts of the saturation variable")
```

• Histograms or Density plots for continuous variables. Are the
distributions Gaussian like?

```{r}
df %>% ggplot(aes(x = R)) + geom_histogram() + labs(title = "Distribution of 'R' (red values in the data)")

df %>% ggplot(aes(x = G)) + geom_histogram() + labs(title = "Distribution of 'G' (green values in the data)")

df %>% ggplot(aes(x = B)) + geom_histogram() + labs(title = "Distribution of 'B' (blue values in the data)")

df %>% ggplot(aes(x = Hue)) + geom_histogram() + labs(title = "Distribution of 'Hue'")
```

Both the R, G, and B values seemed to be a (noisy) skewed distribution,
with a heavy left tail. Hue seems to have certain common values (ie
spikes in the graph), but it looks overall uniform in distribution.

• Condition (group) the continuous variables based on the categorical
variables.

COLORS

```{r}
# Histogram of R grouped by Lightness using faceting
df %>% 
  ggplot(aes(x = R)) +
  geom_histogram(bins = 30, fill = "red", color = "white") +
  facet_wrap(~ Lightness) +
  labs(title = "Distribution of R by Lightness", x = "R value", y = "Count")

df %>% 
  ggplot(aes(x = G)) +
  geom_histogram(bins = 30, fill = "green", color = "white") +
  facet_wrap(~ Lightness) +
  labs(title = "Distribution of G by Lightness", x = "G value", y = "Count")

df %>% 
  ggplot(aes(x = B)) +
  geom_histogram(bins = 30, fill = "blue", color = "white") +
  facet_wrap(~ Lightness) +
  labs(title = "Distribution of B by Lightness", x = "B value", y = "Count")

# Density plot of G grouped by Saturation using faceting
df %>% 
  ggplot(aes(x = R)) +
  geom_density(fill = "tomato", alpha = 0.5) +
  facet_wrap(~ Saturation) +
  labs(title = "Density of R by Saturation", x = "R value", y = "Density")

df %>% 
  ggplot(aes(x = G)) +
  geom_density(fill = "green", alpha = 0.5) +
  facet_wrap(~ Saturation) +
  labs(title = "Density of G by Saturation", x = "G value", y = "Density")

df %>% 
  ggplot(aes(x = B)) +
  geom_density(fill = "blue", alpha = 0.5) +
  facet_wrap(~ Saturation) +
  labs(title = "Density of B by Saturation", x = "B value", y = "Density")
```

• Are there differences in continuous variable distributions and
continuous variable summary statistics based on categorical variable
values?

STRATIFIED BY LIGHTNESS

-   R stratified by Lightness - Noisy but potentially gaussian - Dark,
    deep, saturated Clean distributions (looks to be little noise) -
    light, midtone, pale, soft

-   G stratified by Lightness - potentially gaussian - dark, deep,
    midtone, saturated, soft light might be gaussian, but contains some
    outlier values

-   B stratified by Lightness - Noisy but potentially gaussian -
    midtone, saturated, soft

-   IN AGGREGATE

    -   THE COLORS BY LIGHTNESS SEEM TO HAVE DISTRIBUTIONS THAT WERE
        PREVIOUSLY NOT SEEN IN AGGREGATE

    -   there are many seemingly gaussian distributions that appear when
        you group color by lightness level, and this could be useful in
        making predictions

STRATIFIED BY SATURATION

-   R by Saturation

    -   almost all of the saturation values seem to be normal with an
        extreme left skew "gray" saturation has more in the middle than
        expected if it followed a left skew normal dist

-   G by Saturation

    -   similar to R by saturation, most of the distributions seem to
        have large left tails.

    -   grey seems to be more centered around the mean (\~150) as well

-   B by Saturation

    -   many of these distributions have a bi-modal shape. There is a
        peak around the \~50-100 region, and then another around the
        \~200 region.

    -   grey, again, seems to be more centered around the mean

-   IN AGGREGATE

    -   by saturation, each of the colors individually seem to have the
        most values around the \~200 saturation level. Blue tends to
        have more values in the lower saturation than either red or
        green.

HUE GROUPED BY CATEGORICAL

```{r}
df %>% 
  ggplot(aes(x = Hue)) +
  geom_histogram(bins = 30, fill = "black", color = "white") +
  facet_wrap(~ Lightness) +
  labs(title = "Distribution of hue by Lightness", x = "hue value", y = "Count")

df %>% 
  ggplot(aes(x = Hue)) +
  geom_histogram(bins = 30, fill = "black", color = "white") +
  facet_wrap(~ Saturation) +
  labs(title = "Distribution of hue by Saturation", x = "hue value", y = "Count")
```

-   These values look to be largely uniform distributions with lots of
    noise. It will be hard to gain predictive insight from hue when
    stratified by the Lightness or Saturation value.

• Are there differences in continuous variable distributions and
continuous variable summary statistics based on the binary outcome?

```{r}
df %>% 
  ggplot(aes(x = R, fill = as.factor(outcome))) +
  geom_density(alpha = 0.5) +
  labs(title = "Density of R by Outcome", x = "R value", y = "Density", fill = "Outcome")

df %>% 
  ggplot(aes(x = G, fill = as.factor(outcome))) +
  geom_density(alpha = 0.5) +
  labs(title = "Density of G by Outcome", x = "G value", y = "Density", fill = "Outcome")

df %>% 
  ggplot(aes(x = B, fill = as.factor(outcome))) +
  geom_density(alpha = 0.5) +
  labs(title = "Density of B by Outcome", x = "B value", y = "Density", fill = "Outcome")

df %>% 
  ggplot(aes(x = Hue, fill = as.factor(outcome))) +
  geom_density(alpha = 0.5) +
  labs(title = "Density of Hue by Outcome", x = "Hue value", y = "Density", fill = "Outcome")
```

R BY OUTCOME

-   Both the 0 and 1 outcome seem to have a similar distribution, the 0
    outcomes have a higher density amongst high r values, but the
    general shape and mode of both distributions are similar.

G BY OUTCOME

-   Low values (50-100) of G have a higher density in the 1 outcome.
    Values in the (\~100-160) range have the opposite pattern, higher
    density of 0 than of 1. the upper values seem to be very similar
    when stratified by outcome.

B BY OUTCOME

-   the shape of these distributions are very similar. Blue distribution
    with outcome 0 seems to be bi-modal, with a first peak around 80,
    where outcome 1's stratification seems to be unimodal. both peaks in
    the high range of B are around 210.

HUE BY OUTCOME

-   this plot seems to be the most interesting of them all. Hue by 0
    seems to be inversely correlated with hue by 1 outcomes. This means,
    at least it appears visually, as the density of Hue goes up when
    looking at just 0 outcomes, the density of the hue with 1 outcomes
    goes down. Since the two distributions are very different, looking
    at the relative densities of the two could provide a useful insight
    in predicting outcome with just a hue value. This can be used as
    part of a larger model, but it seems to be a promising idea.

• Visualize the relationships between the continuous inputs, are they
correlated?

```{r}
df %>% ggplot(aes(x = R, y = G)) + geom_point()

df %>% ggplot(aes(x = R, y = B)) + geom_point()

df %>% ggplot(aes(x = B, y = G)) + geom_point()

df %>% ggplot(aes(x = Hue, y = R)) + geom_point()

df %>% ggplot(aes(x = Hue, y = B)) + geom_point()

df %>% ggplot(aes(x = Hue, y = G)) + geom_point()
 
```

Relationships between continuous variables:\

All color relationships have a strong cluster in the top left (ie when
one color is of high value it is common for the other color will be of
high value)

Red and Green: seems to have a certain relationship, but it looks non
linear. Could be quadratic or of another nature.

Red and Blue: seems to have little to no relationship.

Blue and Green: seems to have a linear relationship.

Hue and the colors do not seem to produce a graph that looks more than
noise.

• Visualize the relationships between the continuous outputs (response
and the LOGIT-transformed response, y) with respect to the continuous
INPUTS.

```{r}
dfii <- df %>% 
  mutate(y = boot::logit( (response - 0) / (100 - 0) ) ) %>% 
  select(R, G, B, 
         Lightness, Saturation, Hue, response,
         y)


dfii %>% glimpse()
```

```{r}
dfii %>% ggplot(aes(x = R, y = response)) + geom_point(color = "red") + labs(title = "R VS RESPONSE")
dfii %>% ggplot(aes(x = R, y = y)) + geom_point(color = "red") + labs(title = "R VS LOGIT RESPONSE")
```

Red seems to have a quadratic or other non linear relationship with
logit response. The relationship seems less obvious when compared with
non-transformed response.

```{r}
dfii %>% ggplot(aes(x = R, y = response)) + geom_point(color = "tomato") + facet_wrap(~ Lightness) + labs(title = "R VS RESPONSE BY LIGHTNESS")
dfii %>% ggplot(aes(x = R, y = y)) + geom_point(color = "tomato") + facet_wrap( ~Lightness) + labs(title = "R VS LOGIT RESPONSE BY LIGHTNESS")
```

When grouped by lightness, certain lightness levels (light, pale, soft)
have formed clusters, meaning that for a paint with this lightness
level, there is a relatively small range that both the response and the
color will be in.

```{r}
dfii %>% ggplot(aes(x = R, y = response)) + geom_point(color = "darkred") + facet_wrap(~ Saturation) + labs(title = "R VS RESPONSE BY SATURATION")
dfii %>% ggplot(aes(x = R, y = y)) + geom_point(color = "darkred") + facet_wrap(~ Saturation) + labs(title = "R VS LOGIT RESPONSE BY SATURATION")
```

When grouped by saturation, certain saturation levels show a very linear
relationship between red and logit response (this applies to neutral,
gray, and shaded). The other satuartion levels seem to fit a sort of
tilted parabolic shape.

```{r}
dfii %>% ggplot(aes(x = G, y = response)) + geom_point(color = "green") + labs(title = "G VS RESPONSE")
dfii %>% ggplot(aes(x = G, y = y)) + geom_point(color = "green") + labs(title = "G VS LOGIT RESPONSE")
```

Green vs Logit response seems to look like a very strong linear
relationship.

```{r}
dfii %>% ggplot(aes(x = G, y = response)) + geom_point(color = "lightgreen") + facet_wrap(~ Lightness) + labs(title = "G VS RESPONSE BY LIGHTNESS")
dfii %>% ggplot(aes(x = G, y = y)) + geom_point(color = "lightgreen") + facet_wrap(~ Lightness) + labs(title = "G VS LOGIT RESPONSE BY LIGHTNESS")
```

We can see this partial clustering happening with grouping by lightness
(as we saw in red as well). The ones that seem to cluster most are
(pale, light, and soft).

```{r}
dfii %>% ggplot(aes(x = G, y = response)) + geom_point(color = "darkgreen") + facet_wrap(~ Saturation) + labs(title = "G VS RESPONSE BY SATURATION")
dfii %>% ggplot(aes(x = G, y = y)) + geom_point(color = "darkgreen") + facet_wrap(~ Saturation) + labs(title = "G VS LOGIT RESPONSE BY SATURATION")
```

All of the logit responses vs green look extremely linear when
stratified by saturation. This could be extremely useful in predicting
logit output. The strongest relationships appear to be neutral, gray,
shaded, and subdued.

```{r}
dfii %>% ggplot(aes(x = B, y = response)) + geom_point(color = "steelblue") + labs(title = "B VS RESPONSE")
dfii %>% ggplot(aes(x = B, y = y)) + geom_point(color = "steelblue") + labs(title = "B VS LOGIT RESPONSE")
```

Blue seems to follow a linear relationship with logit response, though
it seems to be relatively weak.

```{r}
dfii %>% ggplot(aes(x = B, y = response)) + geom_point(color = "lightblue") + facet_wrap(~ Lightness) + labs(title = "G VS RESPONSE BY LIGHTNESS")
dfii %>% ggplot(aes(x = B, y = y)) + geom_point(color = "lightblue") + facet_wrap(~ Lightness) + labs(title = "G VS LOGIT RESPONSE BY LIGHTNESS")
```

Again, we can see the clustering when we group by lightness.

```{r}
dfii %>% ggplot(aes(x = B, y = response)) + geom_point(color = "darkblue") + facet_wrap(~ Saturation) + labs(title = "G VS RESPONSE BY SATURATION")
dfii %>% ggplot(aes(x = B, y = y)) + geom_point(color = "darkblue") + facet_wrap(~ Saturation) + labs(title = "G VS LOGIT RESPONSE BY SATURATION")
```

We can also, again, see linear relationships appear when we stratify by
saturation. The strongest appearance of a linear relationship appear to
be neutral, gray, shaded, and subdued saturation levels.

```{r}
dfii %>% ggplot(aes(x = Hue, y = response)) + geom_point(color = "black") + labs(title = "HUE VS RESPONSE")
dfii %>% ggplot(aes(x = Hue, y = y)) + geom_point(color = "black") + labs(title = "HUE VS LOGIT RESPONSE")
```

Hue vs. logit response appears to be almost purely noise. There is no
obvious relationship here.

```{r}
dfii %>% ggplot(aes(x = Hue, y = response)) + geom_point(color = "black") + facet_wrap(~ Lightness) + labs(title = "HUE VS RESPONSE BY LIGHTNESS")
dfii %>% ggplot(aes(x = Hue, y = y)) + geom_point(color = "black") + facet_wrap(~ Lightness) + labs(title = "HUE VS LOGIT RESPONSE BY LIGHTNESS")
```

When stratified by lightness, we can see some almost horizontal ranges
form. This means that regardless of hue (or any other continuous
variable), there is a range of logit responses that the lightness level
will fall under. This suggests that lightness alone, can be useful in
predicting logit response. Some of the lightness levels (notably dark,
deep, and saturated), seem to have higher logit responses are closer to
the middle (\~10-20).

```{r}
dfii %>% ggplot(aes(x = Hue, y = response)) + geom_point(color = "black") + facet_wrap(~ Saturation) + labs(title = "HUE VS RESPONSE BY SATURATION")
dfii %>% ggplot(aes(x = Hue, y = y)) + geom_point(color = "black") + facet_wrap(~ Saturation) + labs(title = "HUE VS LOGIT RESPONSE BY SATURATION")
```

There seems to be little relationship between hue and logit response,
even when stratified by saturation level.

• Can you identify any clear trends? Do the trends depend on the
categorical INPUTS?

Many trends appear when we visualise the relationship between the
continuous variables and the response/logit response. Some stronger
trends appear when you group by categorical inputs.

• How can you visualize the behavior of the binary outcome with respect
to the continuous inputs? How can you visualize the behavior of the
binary outcome with respect to the categorical INPUTS?

The first can be done with a density plot (already done above).

Plotting categorical variables relative to the binary inputs can be done
with a side-by-side bar chart as follows:

```{r}
df %>% 
  ggplot(aes(x = Lightness, fill = as.factor(outcome))) +
  geom_bar(position = "dodge") +
  labs(title = "Counts of Outcome by Lightness",
       x = "Lightness",
       y = "Count",
       fill = "Outcome")

df %>% 
  ggplot(aes(x = Saturation, fill = as.factor(outcome))) +
  geom_bar(position = "dodge") +
  labs(title = "Counts of Outcome by Saturation",
       x = "Saturation",
       y = "Count",
       fill = "Outcome")

```

Lightness seems to have a relatively constant ratio of 0 and 1 outcomes
per each lightness level. However, Saturations eems to have a widely
varying ratio of 0s to 1s depending on the saturation level. This might
be useful in predictions.
