---
title: "Final P3"
author: "Jonah Smith"
date: "2025-04-18"
output: html_document
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(coefplot)
library(caret)
library(rstanarm)
library(pROC)
library(broom)
library(glmnet)
library(nnet)
library(tidyr)
library(bayesplot)
```

```{r}
## function to be used for feature analysis in the models
plot_top_t_coefs <- function(model, top_n = 10, title = "Top Predictors by |t-value|") {
  broom::tidy(model) %>%
    mutate(t_value = estimate / std.error) %>%
    arrange(desc(abs(t_value))) %>%
    slice(1:top_n) %>%
    ggplot(aes(x = reorder(term, abs(t_value)), y = estimate)) +
    geom_col(fill = "steelblue") +
    geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error), width = 0.2) +
    coord_flip() +
    labs(title = title,
         x = "Predictor",
         y = "Coefficient Estimate") +
    theme_minimal()
}
```


```{r, echo=TRUE}
# Load your training CSV into df so that df %>% mutate() will work
df <- read_csv("paint_project_train_data.csv", col_names = TRUE)

# Prepare classification dataset: include only predictors + the binary outcome
dfiii <- df %>%
  mutate(outcome = factor(outcome)) %>%   # ensure outcome is a factor 0/1
  select(R, G, B, Hue, Lightness, Saturation, outcome)

# 1) Intercept‑only
glm0 <- glm(outcome ~ 1,
            data   = dfiii,
            family = binomial)

# 2) Categorical only
glm_cat <- glm(outcome ~ factor(Lightness) + factor(Saturation),
               data   = dfiii,
               family = binomial)

# 3) Continuous only
glm_cont <- glm(outcome ~ R + G + B + Hue,
                data   = dfiii,
                family = binomial)

# 4) All additive
glm_all <- glm(outcome ~ R + G + B + Hue +
                        factor(Lightness) + factor(Saturation),
               data   = dfiii,
               family = binomial)

# 5) Main‑effects × Categorical interactions
glm_int <- glm(outcome ~ (R + G + B + Hue) *
                        (factor(Lightness) + factor(Saturation)),
               data   = dfiii,
               family = binomial)

# 6) Pairwise continuous + categories
glm_pwcat <- glm(outcome ~ (R + G + B + Hue)^2 +
                           factor(Lightness) + factor(Saturation),
                 data   = dfiii,
                 family = binomial)

# 7) Full pairwise × categories
glm_total <- glm(outcome ~ (R + G + B + Hue)^2 *
                            (factor(Lightness) + factor(Saturation)),
                 data   = dfiii,
                 family = binomial)

#MY MODELS
#hue and saturation seem to be the most significant factors in the part 1 analysis for classifying outcome, just include them for a simple model
my_glm_1 <- glm(outcome ~ (Hue) + factor(Saturation),
                data    = dfiii,
                family = binomial)

my_glm_2 <- glm(outcome ~ (Hue) * factor(Saturation),
                data    = dfiii,
                family = binomial)

##consider the circular approach that I used for the regression model in P2. Since it seemed to capture something interesting, we'll try it again here
my_glm_3 <- glm(
  outcome ~ sinpi(Hue/180) + cospi(Hue/180) +
             factor(Saturation) +
             sinpi(Hue/180):factor(Saturation) +
             cospi(Hue/180):factor(Saturation),
  data   = dfiii,
  family = binomial
)

# Summarize all models
###
list(
  intercept_only = summary(glm0),
  categorical    = summary(glm_cat),
  continuous     = summary(glm_cont),
  all_additive   = summary(glm_all),
  interactions   = summary(glm_int),
  pairwise_cat   = summary(glm_pwcat),
  full_model     = summary(glm_total),
  my_glm_1       = summary(my_glm_1),
  my_glm_2       = summary(my_glm_2),
  my_gml_3       = summary(my_glm_3)
)

#glimpse(dfiii)
```
I recieved the warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

This essentially means that the models start to predict so large a probability that it will essentially over or underflow to 0 or 1. I think this is to be expected when we have models with this complexity. 



```{r}
# 1) Build a tibble of true labels + predicted probabilities
prob_df <- tibble(
  true          = dfiii$outcome,
  intercept     = predict(glm0,         type = "response"),
  categorical   = predict(glm_cat,      type = "response"),
  continuous    = predict(glm_cont,     type = "response"),
  all_additive  = predict(glm_all,      type = "response"),
  interactions  = predict(glm_int,      type = "response"),
  pairwise_cat  = predict(glm_pwcat,    type = "response"),
  full_model    = predict(glm_total,    type = "response"),
  my_glm_1      = predict(my_glm_1,     type = "response"),
  my_glm_2      = predict(my_glm_2,     type = "response"),
  my_glm_3      = predict(my_glm_3,     type = "response"),
)

# 2) Compute ROC coordinates for each model
roc_list <- lapply(names(prob_df)[-1], function(m) {
  roc_obj   <- roc(prob_df$true, prob_df[[m]])
  coords_df <- coords(roc_obj, "all",
                     ret = c("specificity","sensitivity"),
                     transpose = FALSE)
  tibble(model       = m,
         specificity = coords_df$specificity,
         sensitivity = coords_df$sensitivity)
})

roc_df <- bind_rows(roc_list) %>%
  mutate(fpr = 1 - specificity)

# 3) Plot them with ggplot2
ggplot(roc_df, aes(x = fpr, y = sensitivity, color = model)) +
  geom_line(size = 1) +
  geom_abline(linetype = "dashed") +
  scale_x_continuous("False Positive Rate (1–Specificity)", limits = c(0,1)) +
  scale_y_continuous("True Positive Rate (Sensitivity)",   limits = c(0,1)) +
  labs(title = "ROC Curves for Logistic Regression Models",
       color = "Model") +
  theme_minimal()

# 4) Rank models by AUC
model_list <- list(
  intercept     = glm0,
  categorical   = glm_cat,
  continuous    = glm_cont,
  all_additive  = glm_all,
  interactions  = glm_int,
  pairwise_cat  = glm_pwcat,
  full_model    = glm_total,
  my_glm_1      = my_glm_1,
  my_glm_2      = my_glm_2,
  my_glm_3      = my_glm_3
)

auc_df <- tibble(
  model = names(model_list),
  AUC   = sapply(model_list, function(m) {
    roc_obj <- roc(dfiii$outcome, predict(m, type="response"))
    as.numeric(auc(roc_obj))
  })
) %>% 
  arrange(desc(AUC))

print(auc_df)

# 5) Plot top‐10 coefficients (by |z|) for the top 3 models
plot_top_z_coefs <- function(mod, top_n = 10, title = "") {
  tidy(mod) %>%
    mutate(z_value = estimate / std.error) %>%
    arrange(desc(abs(z_value))) %>%
    slice(1:top_n) %>%
    ggplot(aes(x = reorder(term, z_value), y = estimate)) +
      geom_col(fill = "steelblue") +
      geom_errorbar(aes(
        ymin = estimate - std.error, 
        ymax = estimate + std.error
      ), width = 0.2) +
      coord_flip() +
      labs(
        title = title,
        x = "Predictor",
        y = "Coefficient (log‑odds)"
      ) +
      theme_minimal()
}

top3 <- auc_df$model[1:3]

for (m in top3) {
  print(plot_top_z_coefs(
    model_list[[m]], 
    top_n = 10, 
    title = paste("Top 10 Coefs by |z| for", m)
  ))
}
```
The factors seem important here as well (similar to the regression model). Some of the colors seemed to be important in terms of their interactions with other features. The only continuous feature that sits in the top 10 alone is R for any of the models. The key takeaway is that the categorical features are consistently significant in the classification model.


I struggled with overfitting of the full model, so the two models I will use for the bayesian approach will be the interaction and the pairwise concatenation.
```{r}
color_scheme_set("blue")

# 1) Fit the two Bayesian GLMs
stan_int <- stan_glm(
  outcome ~ (R + G + B + Hue) *
                           (factor(Lightness) + factor(Saturation)),
  data   = dfiii,
  family = binomial(link = "logit"),
  chains = 4, iter = 2000
)

stan_total <- stan_glm(
  outcome ~ (R + G + B + Hue)^2 *
                            (factor(Lightness) + factor(Saturation)),
  data   = dfiii,
  family = binomial(link = "logit"),
  chains = 4, iter = 2000
)

# 2) Compare AUCs via posterior‐mean probabilities
post_int  <- posterior_linpred(stan_int,  transform = TRUE)
post_total <- posterior_linpred(stan_total, transform = TRUE)
mean_int  <- colMeans(post_int)
mean_total <- colMeans(post_total)

auc_int  <- auc(dfiii$outcome, mean_int)
auc_total <- auc(dfiii$outcome, mean_total)

```

```{r}
auc_int
auc_total
```


```{r}
# 1) pull out 95% posterior intervals
ci_mat <- posterior_interval(stan_int, prob = 0.95)

# 2) keep only those terms whose interval does NOT contain 0
keep_pars <- rownames(ci_mat)[ ci_mat[,1] > 0 | ci_mat[,2] < 0 ]

# 3) plot densities & CIs for just those "nonzero" coefficients
mcmc_areas(
  as.array(stan_int),
  pars = keep_pars,
  prob = 0.95
) +
  ggtitle("95% Posterior Intervals (Excluding CIs Crossing 0)") +
  xlab("Log‑odds Coefficient") +
  ylab("Feature") +
  theme_minimal()

```
```{r}
# 1) pull out 95% posterior intervals
ci_mat <- posterior_interval(stan_total, prob = 0.95)

# 2) keep only those terms whose interval does NOT contain 0
keep_pars <- rownames(ci_mat)[ ci_mat[,1] > 0 | ci_mat[,2] < 0 ]

# 3) plot densities & CIs for just those "nonzero" coefficients
mcmc_areas(
  as.array(stan_total),
  pars = keep_pars,
  prob = 0.95
) +
  ggtitle("95% Posterior Intervals (Excluding CIs Crossing 0)") +
  xlab("Log‑odds Coefficient") +
  ylab("Feature") +
  theme_minimal()

```

Despite the total model having a better AUC value, it seems that there is a heavy risk of this not generalizing when we do a cross validation fit. Only one feature is statistically significant. For this reason, I would say that the interactions model is likely going to be the best in terms of predictive capability (despite lower ROC val).


```{r}
# Step 1: Setup prediction grid
# Fixed values for remaining vars
ref_vals <- dfiii %>%
  summarise(across(c(R, G, B, Lightness), median))

# Choose 5 representative saturation levels
sat_levels <- dfiii %>%
  distinct(Saturation) %>%
  arrange(Saturation) %>%
  pull(Saturation) %>%
  unique() %>%
  sort()

# Pick 5 roughly evenly spaced levels
sat_sample <- sat_levels[round(seq(1, length(sat_levels), length.out = 5))]

# Create prediction grid
pred_grid <- expand_grid(
  Hue        = seq(0, 360, length.out = 100),
  Saturation = sat_sample
) %>%
  mutate(
    R         = ref_vals$R,
    G         = ref_vals$G,
    B         = ref_vals$B,
    Lightness = ref_vals$Lightness
  )

# Step 2: Predict from both Bayesian models
post_int_pred  <- posterior_linpred(stan_int,  newdata = pred_grid, transform = TRUE)
post_total_pred <- posterior_linpred(stan_total, newdata = pred_grid, transform = TRUE)

# Step 3: Summarize into mean & 95% CI
summarize_preds <- function(mat) {
  tibble(
    mean  = colMeans(mat),
    lower = apply(mat, 2, quantile, 0.025),
    upper = apply(mat, 2, quantile, 0.975)
  )
}

pred_grid_int  <- bind_cols(pred_grid, summarize_preds(post_int_pred))  %>% mutate(model = "stan_int")
pred_grid_total <- bind_cols(pred_grid, summarize_preds(post_total_pred)) %>% mutate(model = "stan_total")

plot_df <- bind_rows(pred_grid_int, pred_grid_total)

# Step 4: Plot
ggplot(plot_df, aes(x = Hue, y = mean, color = model, fill = model)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, color = NA) +
  geom_line() +
  facet_wrap(~ Saturation, nrow = 2) +
  labs(
    title = "Predicted Event Probability vs Hue by Saturation",
    x = "Hue",
    y = "Predicted Probability",
    color = "Model",
    fill  = "Model"
  ) +
  theme_minimal()

```
Though we see a very high range of uncertainty for both models, it appears we are capturing the same underlying trend here. The posterior means are generally very close and follow the same trajectory. 

```{r}
df_cl <- df %>%
  mutate(outcome = factor(ifelse(outcome == 1, "yes", "no"))) %>%
  select(R, G, B, Hue, Lightness, Saturation, outcome)

# 1. Resampling config: 5-fold CV
ctrl <- trainControl(
  method = "cv",
  number = 5,
  summaryFunction = twoClassSummary,
  classProbs = TRUE,
  savePredictions = "final"
)

metric <- "ROC"  # optimizing ROC AUC

# 2. Helper function to train a model
train_model <- function(formula, method, tuneLength = 5) {
  train(
    formula,
    data = df_cl,
    method = method,
    metric = metric,
    trControl = ctrl,
    preProcess = c("center", "scale"),
    tuneLength = tuneLength
  )
}

# 3. Train all models
models <- list(
  GLM_Additive  = train_model(outcome ~ R + G + B + Hue + Lightness + Saturation, "glm"),
  GLM_Interact  = train_model(outcome ~ (R + G + B + Hue)^2 + factor(Lightness) + factor(Saturation), "glm"),
  GLM_My_2      = train_model(outcome ~ Hue * factor(Saturation), "glm"),
  GLM_My_3      = train_model(outcome ~ sinpi(Hue/180) + cospi(Hue/180) + factor(Saturation) +
                                            sinpi(Hue/180):factor(Saturation) +
                                            cospi(Hue/180):factor(Saturation), "glm"),
  ElasticNet    = train_model(outcome ~ (R + G + B + Hue)^2 + Lightness + Saturation, "glmnet", tuneLength = 10),
  Neural_Net    = train_model(outcome ~ ., "nnet", tuneLength = 10),
  Random_Forest = train_model(outcome ~ ., "rf", tuneLength = 5),
  Naive_Bayes   = train_model(outcome ~ ., "nb", tuneLength = 5),
  kNN           = train_model(outcome ~ ., "knn", tuneLength = 10)
)

# 4. Compare resampled results
res <- resamples(models)
```

```{r}
summary(res, metric = c("ROC", "Sens", "Spec"))
dotplot(res, metric = "ROC")
```
The random forest model is the best model in maximizing the ROC. This was my chosen performance metric, so it is also the best model overall. 


SAVING THE BEST MODEL FOR P4

```{r}
models$Random_Forest %>% readr::write_rds("rand_forest.rds")
```

```{r}
rand_forest <- readr::read_rds("rand_forest.rds")

glimpse(rand_forest)
```




